---
title: "Cold Pool Index"
author: 
- affiliation: RACE
  description: Research Fish Biologist
  email: Sean.Rohan@noaa.gov
  name: Sean Rohan
output: word_document
fontsize: 12pt
addr: 
  l1: 7600 Sand Point Way NE
  l2: NMFS RACE Division, Groundfish Assessment Program
  l3: Seattle, WA 98115
---

# Cold Pool Index


```{r calc_index, message=FALSE, warning=FALSE, include=FALSE}
library(coldpool)
library(ggrepel)

fig_res <- 300
```

```{r plot_index}
dataset_date <- coldpool:::cold_pool_index$LAST_UPDATE[1]
cpi_sd <- sd(coldpool:::cold_pool_index$AREA_LTE2_KM2)
cpi_mean <- mean(coldpool:::cold_pool_index$AREA_LTE2_KM2)
bt_sd <- sd(coldpool:::cold_pool_index$MEAN_GEAR_TEMPERATURE)
bt_mean <- mean(coldpool:::cold_pool_index$MEAN_GEAR_TEMPERATURE)
sst_sd <- sd(coldpool:::cold_pool_index$MEAN_SURFACE_TEMPERATURE)
sst_mean <- mean(coldpool:::cold_pool_index$MEAN_SURFACE_TEMPERATURE)

cpindex_df <- coldpool:::cold_pool_index %>%
  dplyr::mutate(cpi_dev = (AREA_LTE2_KM2-cpi_mean)/cpi_sd,
                bt_dev = (MEAN_GEAR_TEMPERATURE-bt_mean)/bt_sd,
                sst_dev = (MEAN_SURFACE_TEMPERATURE-sst_mean)/sst_sd,
                cpi_rank = rank(-1*AREA_LTE2_KM2),
                bt_rank = rank(-1*MEAN_GEAR_TEMPERATURE),
                sst_rank = rank(-1*MEAN_SURFACE_TEMPERATURE))

cpa_plot <- ggplot(data = cpindex_df,
             aes(x = YEAR, 
                 y = AREA_LTE2_KM2,
                 label = cpi_rank)) +
  geom_line() +
  geom_point(aes(fill = cut(cpi_dev, c(-3, -2, -1, 1, 2, 3))),
             shape = 21,
             size = rel(3)) +
  ggrepel::geom_text_repel() +
  scale_fill_brewer(name = "Std. dev.", 
                    drop = FALSE, 
                    palette = "RdYlBu",
                    direction = 1) +
  scale_x_continuous(name = "Year") +
  scale_y_continuous(name = expression('Cold Pool Area'~(km^2))) +
  theme_bw()

bt_mean_plot <- ggplot(data = cpindex_df,
             aes(x = YEAR, 
                 y = MEAN_GEAR_TEMPERATURE,
                 label = bt_rank)) +
  geom_line() +
  geom_point(aes(fill = cut(bt_dev, c(-3, -2, -1, 1, 2, 3))),
             shape = 21,
             size = rel(3)) +
    ggrepel::geom_text_repel() +
  scale_fill_brewer(name = "Std. dev.", 
                    drop = FALSE, 
                    palette = "RdYlBu",
                    direction = -1) +
  scale_x_continuous(name = "Year") +
  scale_y_continuous(name = expression('Mean Bottom Temperature'~(degree*C))) +
  theme_bw()

sst_mean_plot <- ggplot(data = cpindex_df,
             aes(x = YEAR, 
                 y = MEAN_SURFACE_TEMPERATURE,
                 label = sst_rank)) +
  geom_line() +
  geom_point(aes(fill = cut(sst_dev, c(-3, -2, -1, 1, 2, 3))),
             shape = 21,
             size = rel(3)) +
    ggrepel::geom_text_repel() +
  scale_fill_brewer(name = "Std. dev.", 
                    drop = FALSE, 
                    palette = "RdYlBu",
                    direction = -1) +
  scale_x_continuous(name = "Year") +
  scale_y_continuous(name = expression('Mean Surface Temperature'~(degree*C))) +
  theme_bw()

png(file = here::here("plots", "cold_pool_area.png"), width = 6, height = 4, res = fig_res, units = "in")
print(cpa_plot)
dev.off()

png(file = here::here("plots", "bottom_temperature_mean.png"), width = 6, height = 4, res = fig_res, units = "in")
print(bt_mean_plot)
dev.off()

png(file = here::here("plots", "surface_temperature_mean.png"), width = 6, height = 4, res = fig_res, units = "in")
print(sst_mean_plot)
dev.off()

coldpool:::cold_pool_index$YEAR[order(coldpool:::cold_pool_index$AREA_LTE2_KM2)]
coldpool:::cold_pool_index$YEAR[order(-1*coldpool:::cold_pool_index$MEAN_GEAR_TEMPERATURE)]
coldpool:::cold_pool_index$YEAR[order(-1*coldpool:::cold_pool_index$MEAN_SURFACE_TEMPERATURE)]
```

Contributed by Sean Rohan^1^ and Lewis Barnett^1^   
^1^ Resource Assessment and Conservation Engineering Division, Alaska Fisheries Science Center, National Marine Fisheries Service, NOAA  
**Contact**: sean.rohan@noaa.gov
**Last updated**:  October 2021

**Description of Indicator**: The cold pool is defined as the area of the southeastern Bering Sea continental shelf with bottom temperature $\leq$ 2&deg;C, in square kilometers.

**Methodological Changes**:

**Status and Trends**:

**Implications**:

**Research Priorities**: