---
title: "Cold Pool Index"
author: 
- affiliation: RACE
  description: Research Fish Biologist
  email: Sean.Rohan@noaa.gov
  name: Sean Rohan
output: word_document
fontsize: 12pt
addr: 
  l1: 7600 Sand Point Way NE
  l2: NMFS RACE Division, Groundfish Assessment Program
  l3: Seattle, WA 98115
---

# Cold Pool Index


```{r calc_index, message=FALSE, warning=FALSE, include=FALSE}
library(coldpool)

# Load cold pool index data
cpindex_df <- coldpool:::cold_pool_index
dataset_date <- cpindex_df$LAST_UPDATE[1]

```

```{r plot_index}
cpindex_df <- coldpool:::cold_pool_index
cpi_sd <- sd(cpindex_df$AREA_LTE2_KM2)
cpi_mean <- mean(cpindex_df$AREA_LTE2_KM2)
cpindex_df <- cpindex_df %>%
  dplyr::mutate(dev = (AREA_LTE2_KM2-cpi_mean)/cpi_sd)

ggplot(data = cpindex_df,
             aes(x = YEAR, 
                 y = AREA_LTE2_KM2)) +
  geom_line() +
  geom_point(aes(fill = cut(dev, c(-3, -2, -1, 1, 2, 3))),
             shape = 21,
             size = rel(3)) +
  scale_fill_brewer(name = "Std. dev.", 
                    drop = FALSE, 
                    palette = "RdYlBu") +
  scale_x_continuous(name = "Year") +
  scale_y_continuous(name = expression('Cold Pool Area'~(km^2))) +
  theme_bw()
```

Contributed by Sean Rohan^1^ and Lewis Barnett^1^   
^1^ Resource Assessment and Conservation Engineering Division, Alaska Fisheries Science Center, National Marine Fisheries Service, NOAA  
**Contact**: sean.rohan@noaa.gov
**Last updated**:  October 2021

**Description of Indicator**: The cold pool is defined as the area of the southeastern Bering Sea continental shelf with bottom temperature $\leq$ 2&deg;C, in square kilometers.

**Methodological Changes**:

**Status and Trends**:

**Implications**:

**Research Priorities**: