---
title: "Compare interpolation methods"
author: "Sean Rohan and Lewis Barnett"
date: "9/24/2021"
output: word_document
---

# Get data

Get temperature data from racebase and survey tables.

```{r setup, warning=FALSE, message=FALSE, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(coldpool)

# Plotting ----
fig_res <- 300

# CRS
proj_crs <- coldpool:::ebs_proj_crs

# Retrieve area-weighted surface and bottom temperatures
channel <- coldpool:::get_connected(schema = "AFSC")

# PlusNW strata
area_weighted_plusnw_df <- RODBC::sqlQuery(channel = channel,
                                           query = "select * from HAEHNR.BTTEMP_EBS_PLUSNW_LT200M") |>
  saveRDS(file = here::here("data", "ebs_bttemp_plus_nw_lt200m.rds"))

# Standard strata
area_weighted_standard_df <- RODBC::sqlQuery(channel = channel,
                                             query = "select * from HAEHNR.EBS_BTTEMP_LT200M_STANDARD") |>
  saveRDS(file = here::here("data", "ebs_bttemp_standard_lt200m.rds"))

# Get temperature data
# Writes: 
# -- /data/ebs_nbs_temperature_full_area.csv
# -- /data/index_hauls_temperature_data.csv
coldpool::get_data(channel = channel)
```

# Conduct cross-validation on gear temperature

Perform leave-one-out cross validation (LOOCV) on gear temperature (i.e., bottom temperature) and write results to csv files.

```{r loocv, warning=FALSE, message=FALSE, results='hide'}
# Leave one out cross-validation to compare interpolation methods ----
# Writes:
# -- /plots/RSPE_violin_GEAR_TEMPERATURE_[n].png: Plots of root square prediction error for interpolation methods.
# -- /output/[date]_rmse_loocv_GEAR_TEMPERATURE_[year].csv: Results of leave-one-out cross validation.
coldpool::loocv_gear_temp(temp_data_path =  here::here("data", "index_hauls_temperature_data.csv"),  # update file name to specify data file 
                          proj_crs = proj_crs,
                          interp_variable = "surface_temperature")

coldpool::loocv_gear_temp(temp_data_path =  here::here("data", "index_hauls_temperature_data.csv"),  # update file name to specify data file 
                          proj_crs = proj_crs,
                          interp_variable = "gear_temperature")

# Calculate cold pool area using interpolation methods ----
# Writes:
# -- /output/raster/[method]_[year]_gear_temperature.tif: GeoTIFF raster files for each interpolation method and year
cpa_df <- interpolation_wrapper(temp_data_path = here::here("data", "index_hauls_temperature_data.csv"),
                                proj_crs = coldpool:::ebs_proj_crs,
                                cell_resolution = 5000, # In meters
                                select_years = 1982:2021, 
                                interp_variable = "gear_temperature",
                                select_region = "sebs")
```

# Compare interpolation methods

Create plots and table output showing results of LOOCV.

```{r compare_cpa}
# Make summary tables
make_loocv_summary(sel_paths = list.files("./output/loocv", full.names = TRUE),
                   interp_variable = "surface_temperature")
make_loocv_summary(sel_paths = list.files("./output/loocv", full.names = TRUE),
                   interp_variable = "gear_temperature")

# Set up method labels 
method_labels_df <- data.frame(method = c("BES",
                                          "CIR",
                                          "EXP",
                                          "IDW",
                                          "IDW_NMAX4",
                                          "MAT",
                                          "NN",
                                          "SPH",
                                          "STE",
                                          "TPS",
                                          "OLD IDW"),
                               label = c("OK-BES",
                                         "OK-CIR",
                                         "OK-EXP",
                                         "IDW",
                                         "IDW (Max 4)",
                                         "OK-MAT",
                                         "NN",
                                         "OK-SPH",
                                         "OK-STE",
                                         "TPS",
                                         "Historical"))

# Set up Manual plot colors 
color_vec <- c("OK-BES" = "#4E79A7",
               "OK-CIR" = "#A0CBE8",
               "OK-EXP" = "#F28E2B",
               "IDW" = "#B6992D",
               "IDW (Max 4)" = "#F1CE63",
               "OK-MAT" = "#8CD17D",
               "NN" = "#499894",
               "OK-SPH" = "#FFBE7D",
               "OK-STE" = "#59A14F",
               "TPS" = "#E15759",
               "Historical" = "black")

# Load cold pool calculations from Bob Lauth ----
lauth_cpa_df <- coldpool:::cpa_pre2021 %>%
  dplyr::select(YEAR, 
                AREA_SUM_KM2_LTE2) %>%
  dplyr::rename(year = YEAR,
                `Old IDW` = AREA_SUM_KM2_LTE2)

annual_cpa_df <- read.csv(file = here::here("output", 
                                            "estimated_cpa", 
                                            "cpa_out.csv"),
                          stringsAsFactors = FALSE) %>%
  dplyr::filter(year > 1981)

# Select cold pool area calculations for less than or equal to to 2 degrees
annual_cpa_df <- annual_cpa_df[,c(1, which(stringr::str_detect(names(annual_cpa_df), "_lte2")))]

combined_cpa_df <- annual_cpa_df %>% 
  reshape2::melt(id.vars = "year") %>%
  dplyr::bind_rows(lauth_cpa_df %>% 
                     reshape2::melt(id.vars = "year")) %>%
  dplyr::mutate(method = stringr::str_remove(variable, "_lte2") %>% 
                  toupper()) %>%
  dplyr::inner_join(method_labels_df)

png(file = here::here("plots", "cpa_by_year.png"), width = 169, height = 81, units = "mm", res = fig_res)
print(ggplot() +
  geom_point(data = combined_cpa_df,
             aes(x = year, 
                 y = value, 
                 color = label)) +
  geom_line(data = combined_cpa_df,
            aes(x = year, 
                y = value, 
                color = label)) +
    geom_point(data = combined_cpa_df %>%
                 dplyr::filter(method == "OLD IDW"),
               aes(x = year, 
                   y = value)) +
    geom_line(data = combined_cpa_df %>%
                dplyr::filter(method == "OLD IDW"),
              aes(x = year, 
                  y = value)) +
  scale_x_continuous(name = "Year") +
  scale_color_manual(name = "Method", 
                     values = color_vec) +
  scale_y_continuous(name = expression("Cold Pool Area"~(km^2))) +
  theme_bw() +
    theme(legend.title = element_blank()))
dev.off()

cpa_change_df <- annual_cpa_df %>% 
  reshape2::melt(id.vars = "year") %>%
  dplyr::inner_join(lauth_cpa_df) %>%
  dplyr::mutate(rel_diff = 100*(value - `Old IDW`)/`Old IDW`,
                abs_diff = value - `Old IDW`) %>%
  dplyr::mutate(method = stringr::str_remove(variable, "_lte2") %>% 
                  toupper()) %>%
  dplyr::inner_join(method_labels_df)

png(file = here::here("plots", "cpa_by_year_methods_rel_diff.png"), width = 169, height = 81, units = "mm", res = fig_res)
print(ggplot(data = cpa_change_df,
       aes(x = year,
           y = rel_diff,
           color = label)) + 
  geom_point() + 
  geom_line() +
  scale_y_continuous(name = "Relative Difference (%)") +
  scale_x_continuous(name = "Year") +
  scale_color_manual(name = "Method",
                     values = color_vec) +
  theme_bw() +
  theme()
)
dev.off()

png(file = here::here("plots", "cpa_by_year_methods_abs_diff.png"), width = 169, height = 81, units = "mm", res = fig_res)
print(ggplot(data = cpa_change_df,
       aes(x = year,
           y = abs_diff,
           color = label)) + 
  geom_point() + 
  geom_line() +
  scale_y_continuous(name = expression("Absolute Difference"~(km^2))) +
  scale_x_continuous(name = "Year") +
  scale_color_manual(name = "Method",
                     values = color_vec) +
  theme_bw() +
  theme()
)
dev.off()

cpa_change_df %>%
  dplyr::group_by(variable) %>%
  dplyr::summarise(mean_abs_diff = mean(abs_diff),
                   mean_rel_diff = mean (rel_diff),
                   median_rel_diff = median(rel_diff))

png(file = here::here("plots", "cpa_combined_methods_rel_diff.png"), width = 169, height = 81, units = "mm", res = fig_res)
print(ggplot(data = cpa_change_df, 
       aes(x = label,
           y = rel_diff)) + 
  geom_hline(yintercept = 0) +
  geom_boxplot() +
  scale_y_continuous(name = "Relative Difference (%)") +
  scale_x_discrete(name = "Method") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90,
                                   vjust = 0.5,
                                   hjust = 1))
)
dev.off()

png(file = here::here("plots", "cpa_combined_methods_abs_diff.png"), width = 169, height = 81, units = "mm", res = fig_res)
print(ggplot(data = cpa_change_df, 
       aes(x = label,
           y = abs_diff)) + 
  geom_hline(yintercept = 0) +
  geom_boxplot() +
  scale_y_continuous(name = expression("Absolute Difference"~(km^2))) +
  scale_x_discrete(name = "Method") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90,
                                   vjust = 0.5,
                                   hjust = 1))
)
dev.off()
```

# Compare mean bottom and surface temperatures



```{r meantemp}

# mean_temp_df <- data.frame(YEAR = c(1982:2019, 2021),
#                            Mean = numeric(length = 1L))
# 
# sebs_layers <- akgfmaps::get_base_layers(select.region = "sebs", set.crs = coldpool:::ebs_proj_crs)
# 
# for(i in 1:nrow(mean_temp_df)) {
#   mean_temp_ste <- raster::raster(here::here("output", "raster", "sebs", "gear_temperature", paste0("sebs_ste_", mean_temp_df$YEAR[i], "_gear_temperature.tif"))) %>%
#     akgfmaps::rasterize_and_mask(amask = sebs_layers$survey.area)
#   mean_temp_df$Mean[i] <- mean(mean_temp_ste@data@values, na.rm = TRUE)
#   print(plot(mean_temp_ste, main = mean_temp_df$YEAR[i]))
# }

compare_temp_df <- readRDS(file = here::here("data", "ebs_bttemp_plus_nw_lt200m.rds")) %>%
  dplyr::select(YEAR, AVGBSBT, AVGBSST) %>%
  dplyr::rename(`BT[aw]` = AVGBSBT,
                `SST[aw]` = AVGBSST) %>%
  dplyr::full_join(coldpool:::cold_pool_index) |>
  dplyr::select(-AREA_LTE2_KM2, -AREA_LTE1_KM2, -AREA_LTE0_KM2, -AREA_LTEMINUS1_KM2) %>%
  dplyr::rename(`SST[int]` = MEAN_SURFACE_TEMPERATURE,
                `BT[int]` = MEAN_GEAR_TEMPERATURE)


ggplot(data = compare_temp_df %>%
         tidyr::pivot_longer(cols = c(`BT[aw]`, `SST[aw]`, `SST[int]`, `BT[int]`)),
       aes(x = YEAR, 
           y = value,
           color = name,
           shape = name)) +
  geom_line() +
  geom_point() +
  scale_color_colorblind(labels = scales::parse_format()) +
  scale_shape(labels = scales::parse_format()) +
  scale_y_continuous(name = expression(Temperature~(degree*C))) +
  scale_x_continuous(name = "Year") +
  theme_bw() +
  theme(legend.title = element_blank(),
        legend.text = element_text(hjust = 0))



ggplot() +
  geom_point(data = compare_temp_df %>%
               dplyr::mutate(AREA_WEIGHTED_BT - INTERPOLATED_BT),
             aes(x = YEAR, 
                 y = Mean-Historical)) +
  scale_color_colorblind()
```

# Mean bottom and surface temperature from interpolation versus stratum-weighted method

Historically, mean bottom and surface temperatures were calculated using a mean stratum-weighting approach, where temperature data from survey stations were averaged within strata then weighted in proportion to stratum area within the survey domain. However, this approach does not account for the spatial structure of samples within strata and leads to variation in spatial coverage of samples within strata among years due to instances where equipment malfunctions led to spatially autocorrelated gaps in coverage.

Calculating mean temperature from spatially interpolated rasters of bottom and surface temperatures better accounts for the spatial structure of samples and helps account for temperature at locations where data are missing.

```{r}

area_vs_interp_bt_df <- readRDS(file = here::here("data", "ebs_bttemp_standard_lt200m.rds")) %>% 
  dplyr::select(YEAR, AVGBSBT) %>%
  dplyr::rename(`BT Standard`  = AVGBSBT) %>%
  dplyr::full_join(readRDS(file = here::here("data", "ebs_bttemp_plus_nw_lt200m.rds")) %>%
                       dplyr::select(YEAR, AVGBSBT) %>%
  dplyr::rename(`BT PlusNW`  = AVGBSBT)) %>%
  dplyr::mutate(method = "Area Weighted") %>%
  dplyr::full_join(coldpool:::cold_pool_index %>%
                     dplyr::select(YEAR, 
                                   MEAN_GEAR_TEMPERATURE) %>%
                     dplyr::rename(`BT PlusNW` = MEAN_GEAR_TEMPERATURE) %>%
                     dplyr::mutate(method = "Interpolated")) %>%
  reshape2::melt(id.vars = c("YEAR", "method"))

area_vs_interp_sst_df <- readRDS(file = here::here("data", "ebs_bttemp_standard_lt200m.rds")) %>% 
  dplyr::select(YEAR, AVGBSST) %>%
  dplyr::rename(`SST Standard`  = AVGBSST) %>%
  dplyr::full_join(readRDS(file = here::here("data", "ebs_bttemp_plus_nw_lt200m.rds")) %>%
                       dplyr::select(YEAR, AVGBSST) %>%
  dplyr::rename(`SST PlusNW`  = AVGBSST)) %>%
  dplyr::mutate(method = "Area Weighted") %>%
  dplyr::full_join(coldpool:::cold_pool_index %>%
                     dplyr::select(YEAR, 
                                   MEAN_SURFACE_TEMPERATURE) %>%
                     dplyr::rename(`SST PlusNW` = MEAN_SURFACE_TEMPERATURE) %>%
                     dplyr::mutate(method = "Interpolated")) %>%
  reshape2::melt(id.vars = c("YEAR", "method"))

ggplot() +
  geom_histogram(data = temperature_df,
             aes(x = YEAR)) +
  scale_x_continuous(name = "Year") +
  scale_y_continuous(name = "Frequency")

ggplot(data = area_vs_interp_bt_df,
       aes(x = YEAR, 
           y = value,
           shape = variable,
           color = method)) +
  geom_point() +
  geom_line() +
  scale_color_colorblind(name = "Method") +
  scale_shape(name = "Area") +
  scale_x_continuous(name = "Year") +
  scale_y_continuous(name = expression("Mean Bottom Temperature"~(degree*C))) +
  theme(legend.title = element_blank())
  
ggplot(data = area_vs_interp_sst_df,
       aes(x = YEAR, 
           y = value,
           shape = variable,
           color = method)) +
  geom_point() +
  geom_line() +
  scale_color_colorblind(name = "Method") +
  scale_shape(name = "Area") +
  scale_x_continuous(name = "Year") +
  scale_y_continuous(name = expression("Mean Surface Temperature"~(degree*C)))
  theme()

```
